### 2020-05-25 README.Rmd
trivial change

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 16/2,
  fig.height = 9/2,
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

# forestecology

[![Travis Build Status](https://travis-ci.org/rudeboybert/forestecology.svg?branch=master)](https://travis-ci.org/rudeboybert/forestecology)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/forestecology)](https://cran.r-project.org/package=forestecology)

@@

You can install the released version of forestecology from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("forestecology")
```

@@

```{r, eval = FALSE}
# install.packages("remotes")
# remotes::install_github("rudeboybert/forestecology")
```

## Example analysis

We present an example analysis using:

* [Michigan Big Woods](https://doi.org/10.7302/wx55-kt18) research plot data
* [Smithsonian Conservation Biology Institute (SCBI) ForestGEO](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_main_census) research plot data

```{r}
library(tidyverse)
library(forestecology)
library(snakecase)
library(skimr)
library(sf)
library(sfheaders)
# devtools::install_github("rvalavi/blockCV")
library(blockCV)
library(tictoc)
```

### Load & preprocess data

We first load and preprocess the data corresponding to the two census reflecting the time period over which we consider growth in dbh. We load them into R as "tibble" data frames thereby ensuring a standardized input/output format that can be used across all `tidyverse` packages @tidyverse. Furthermore, we ensure that the different variables have the correct names, types (`dbl`, `data`, `factor`).

**Big Woods**:

```{r}
# Read in Big Woods census data from 2008 & 2014
bw_2008 <- 
  "https://deepblue.lib.umich.edu/data/downloads/z603qx485" %>% 
  read_delim(delim = "\t") %>% 
  mutate(spcode = to_any_case(spcode)) %>%
  select(
    treeID = treeid, stemID = stemtag, sp = spcode, quadrat, gx, gy, dbh, 
    date, codes
  )
bw_2014 <- 
  "https://deepblue.lib.umich.edu/data/downloads/1831ck00f" %>% 
  read_delim(delim = "\t") %>% 
  mutate(spcode = to_any_case(spcode)) %>%
  select(
    treeID = treeid, stemID = stemtag, sp = spcode, quadrat, gx, gy, dbh, 
    date, codes
  )

# Read in bigwoods grouping classification data
bw_species <- 
  "https://deepblue.lib.umich.edu/data/downloads/000000086" %>% 
  read_delim(delim = "\t") %>% 
  # convert all to snake case:
  mutate_at(c("species", "genus", "family", "idlevel", "spcode"), to_any_case) %>% 
  # join trait group
  left_join(families, by = c("spcode", "family")) %>% 
  mutate(
    sp = str_sub(genus, 1, 2), 
    sp = str_c(sp, str_sub(species, 1, 2)),
    sp = tolower(sp),
    latin = str_c(genus, species, sep = " "),
    latin = to_any_case(latin)
  ) %>% 
  select(sp = spcode, genus, species, latin, family, trait_group)

bw_2008 <- bw_2008 %>%
  left_join(bw_species,by='sp')

```

**SCBI**:

```{r}
scbi_2013 <- 
  "https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/census-csv-files/scbi.stem2.csv" %>% 
  read_csv() %>% 
  select(
    treeID, stemID, sp, quadrat, gx, gy, dbh, 
    date = ExactDate, codes, status
  )

scbi_2018 <-
  "https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/census-csv-files/scbi.stem3.csv" %>% 
  read_csv() %>% 
  select(
    treeID, stemID, sp, quadrat, gx, gy, dbh, 
    date = ExactDate, codes, status
  ) %>% 
  mutate(dbh = as.numeric(dbh))
```

### Compute growth

We then combine the two census data frames into a single data frame  that now includes a numerical variable `growth` reflecting the average annual growth of dbh in cm. Furthermore, variables that (in theory) remain unchanged between censuses appear only once, such as `gx`, `gy`, species-related variables. Variables that should change between censuses are tagged with `1/2` indicating earlier/later, such as `dbh1/dbh2`, `codes1/codes2`.

The resulting data frames are named with some variation of `growth_df`.

**Big Woods**:

```{r}
census_df1 <- bw_2008
# we need to filter out the resprouts
census_df2 <- bw_2014 %>% 
  filter(!str_detect(codes, 'R'))
id <- "treeID"

bw_growth_df <- 
  # Merge both censuses and compute growth:
  compute_growth(census_df1, census_df2, id) %>% 
  mutate(sp = to_any_case(sp))
```

**SCBI**:

```{r}
census_df1 <- scbi_2013
census_df2 <- scbi_2018
id <- "stemID"

scbi_growth_df <- 
  # Merge both censuses and compute growth:
  compute_growth(census_df1, census_df2, id) %>%
  # they are mesuaring in mm while we are measuring in cm!!!
  mutate(growth = growth/10)
```

```{r}
growth_df <- bind_rows(
  bw_growth_df %>% select(growth) %>% mutate(site = "bw"),
  scbi_growth_df %>% select(growth) %>% mutate(site = "scbi")
)
ggplot(growth_df, aes(x = growth, y = ..density.., fill = site)) +
  geom_histogram(alpha = 0.5, position = "identity", binwidth = 0.05) +
  labs(x = "Average annual growth in dbh (cm per yr)") +
  coord_cartesian(xlim = c(-0.5, 1))
```

### Add spatial information

We now add spatial information to the `growth_df` data frame. Specifically:

1. In order to control for study region "edge effects" (cite Waller?), we need a function that adds "buffers" of trees. In our case, since our model of interspecies competition relies on a spatial definition of who is a "neighboring competitor" and certain explanatory varibles such as biomass are cummulative, we need to ensure that all trees being modeled are not biased to have different neighbor structure, in particular the trees at the boundary of study regions.
1. Assign each tree to a "fold" for cross-validation purposes. Conventional cross-validation schemes assign observations to folds by resampling individual observations at random. However, underlying this scheme is an assumption that the observations are independent. In the case of forest census data, observations exhibit spatial autocorrelation, and thus this dependence must be incorporated in our resampling scheme @roberts2017 @pohjankukka2017. Packages that have implemented spatial cross-validation include @valavi2019

Thus, before we can incorporate the above information to `growth_df`, we need to define two constants:

```{r}
cv_fold_size <- 100
max_dist <- 7.5
```

#### Defining buffers

For a focal tree of interest, the `max_dist` of `r max_dist` meters acts as a radius defining a neighborhood within which all trees are considered competitors. In other words, all trees within `r max_dist` meters of the focal tree are considered competitor trees. Other studies have estimated this distance; we used `r max_dist` meters as an average of estimated values [@canham2004, @canham2006, @uriarte2004, @tatsumi2013]. Following Tobler's first law of geography that "everything is related to everything else, but near things are more related than distant things." @tobler1970firstlawofgeo, we assume that the degree of spatial autocorrelation is inversely-related to distance. However, we further assume that once trees are more than `r max_dist` meters apart, this autocorrelation is negligeable.

**Example of buffer at work**:

```{r}
# Boundary polygon
square_boundary <- tibble(
  x = c(0,0,1,1),
  y = c(0,1,1,0)
) %>% 
  sf_polygon()

# Buffer polygon
square_buffer <- square_boundary %>% 
  st_buffer(dist = -0.1)

ggplot() +
  geom_sf(data = square_boundary) +
  geom_sf(data = square_buffer, col="red") 
```

**Big Woods**:

```{r}
# Bigwoods study region boundary polygon
bw_boundary <- bigwoods_study_region %>% 
  sf_polygon()

# Buffer polygon
bw_buffer <- bw_boundary %>%
  st_buffer(dist = -max_dist)

# Convert data frame to sf object
bw_growth_df <- bw_growth_df %>% 
  st_as_sf(coords = c("gx", "gy"))

# ID which points are in buffer and which are not
buffer_index <- !st_intersects(bw_growth_df, bw_buffer, sparse = FALSE)
bw_growth_df <- bw_growth_df %>% 
  mutate(buffer = as.vector(buffer_index))

# Plot
ggplot() +  
  geom_sf(data = bw_boundary) +
  geom_sf(data = bw_buffer, col="red") +
  # Only random sample of 1000 trees:
  geom_sf(data = bw_growth_df %>% sample_n(1000), aes(col=buffer), size = 0.5)
```

**SCBI**:

```{r}
# Bigwoods study region boundary polygon
scbi_boundary <- tibble(x = c(0,400,400,0,0), y = c(0,0,640,640,0)) %>% 
  sf_polygon()

# Buffer polygon
scbi_buffer <- scbi_boundary %>%
  st_buffer(dist = -max_dist)

# Convert data frame to sf object
scbi_growth_df <- scbi_growth_df %>% 
  st_as_sf(coords = c("gx", "gy"))

# ID which points are in buffer and which are not
buffer_index <- !st_intersects(scbi_growth_df, scbi_buffer, sparse = FALSE)
scbi_growth_df <- scbi_growth_df %>% 
  mutate(buffer = as.vector(buffer_index))

# Plot
ggplot() +
  geom_sf(data = scbi_boundary) +
  geom_sf(data = scbi_buffer, col="red") +
  # Only random sample of 1000 trees:
  geom_sf(data = scbi_growth_df %>% sample_n(1000), aes(col=buffer), size = 0.5)
```

#### Defining spatial cross-validation folds

We use the [`blockCV`](https://github.com/rvalavi/blockCV) package to define the spatial grid, whose elements will act as the folds in our leave-one-out (by "one" we meen "one grid block") cross-validation scheme.

**Big Woods**:

```{r}
set.seed(76)
bw_cv_grid <- spatialBlock(
  speciesData = bw_growth_df, theRange = 100, verbose = FALSE,
  # Some guess work in figuring this out:
  k = 28, xOffset = 0.5, yOffset = 0
)

# Add foldID to data
bw_growth_df <- bw_growth_df %>% 
  mutate(
    foldID = bw_cv_grid$foldID
  )

# Visualize grid. Why does fold 19 repeat?
bw_cv_grid$plots +
  geom_sf(data = bw_growth_df %>% sample_frac(0.2), aes(col=factor(foldID)), size = 0.1)

# Remove weird folds with no trees in them from viz above
bw_growth_df <- bw_growth_df %>%
  filter(!foldID %in% c(19, 23, 21, 17, 8, 19))
```

**SCBI**:

```{r}
scbi_cv_grid <- spatialBlock(
  speciesData = scbi_growth_df, theRange = 100, k = 28, yOffset = 0.9999, verbose = FALSE
)

# Add foldID to data
scbi_growth_df <- scbi_growth_df %>% 
  mutate(
    foldID = scbi_cv_grid$foldID
  )

# Visualize grid
scbi_cv_grid$plots +
  geom_sf(data = scbi_growth_df, aes(col=factor(foldID)), size = 0.1)
```

### Model specification

Next we specify the growth model we want. We will model growth as a function of tree identity, size (in dbh), and the identity and size (in dbh) of its neighbors. The model can be run on different grouping of individuals (e.g., based on species or trait groupings). It can also be run with different notions of competition:

* `model_number = 1`: No competition growth only depends on focal tree `dbh` and grouping.
* `model_number = 2`: Competition but identity of competitor does not matter, just sum of competitior basal area.
* `model_number = 3`: Full model with competition including competitor identity.

**Big Woods**:

```{r}
bw_specs <- bw_growth_df %>% 
  get_model_specs(model_number = 3, species_notion = 'trait_group')
bw_specs
```

Next we create the `focal_vs_comp` data frame which connects each tree in `growth_df` to the trees in its competitive neighborhood range (as defined by `max_dist`). So for example, if `growth_df` consisted of two focal trees with two and three neighbors respectively, `focal_vs_comp` would consist of 5 rows.

This requires the `growth_df` data frame, `max_dist` scalar defining competitive range, `cv_fold_size` defining the size of the spatial cross-validation blocks, and `model_specs` as inputs.

```{r}
focal_vs_comp_bw <- bw_growth_df %>% 
  create_focal_vs_comp(max_dist, model_specs = bw_specs, cv_grid = bw_cv_grid, id = "treeID")
```

```{r}
glimpse(focal_vs_comp_bw)
```

**SCBI**:

```{r}
scbi_specs <- scbi_growth_df %>% 
  get_model_specs(model_number = 3, species_notion = 'sp')
scbi_specs
```

```{r}
focal_vs_comp_scbi <- scbi_growth_df %>% 
  create_focal_vs_comp(max_dist, model_specs = scbi_specs, cv_grid = scbi_cv_grid, id = "stemID")
```

```{r}
glimpse(focal_vs_comp_scbi)
```

### Model fit and prediction

**Big Woods**:

Now we are ready to fit the competition model with `fit_bayesian_model` this takes two inputs: `focal_vs_comp` the data frame with focal trees connected to their competitors and `model_specs` which specifies the notion of competition.

```{r}
bw_fit_model <- focal_vs_comp_bw %>% 
  fit_bayesian_model(model_specs = bw_specs)
```

This output has the posterior parameters for the specified competition model. If `model_number = 1` this includes just the beta values for how focal tree DBH and identity affect growth. If `model_number = 2` it also includes a vector of lambdas for how total competitor basal area affect individuals of each group. If `model_number = 3` then lambda is a matrix which gives how individuals of each focal group respond to the competition of each competitor group. This `posterior_param` output can be used to get predicted growths for each individual (with `predict_bayesain_model`) to test how well the model performs. Or this `posterior_param` output can be plots (either the betas or lambdas) to understand what controls individual growth.

```{r}
bw_growth_df <- focal_vs_comp_bw %>% 
  predict_bayesian_model(model_specs = bw_specs, posterior_param = bw_fit_model) %>% 
  right_join(bw_growth_df, by = c("focal_ID" = "treeID"))

bw_growth_df <- bw_growth_df %>% 
  st_as_sf()

# Observed vs predicted growth  
ggplot(bw_growth_df, aes(x = growth, y = growth_hat)) +
  geom_point(size = 0.5, color = rgb(0, 0, 0, 0.25)) +
  stat_smooth(method = 'lm') +
  geom_abline(slope = 1, intercept = 0) +
  coord_fixed() + 
  labs(
    x = "Observed growth in dbh", y = "Predicted growth in dbh", 
    title = "Predicted vs Observed Growth"
  )

reslab <- expression(paste('Residual (cm ',y^{-1},')'))
bw_growth_df %>% 
  # Need to investigate missingness
  filter(!is.na(growth_hat)) %>% 
  mutate(
    error = growth - growth_hat,
    error_bin = cut_number(error, n = 5), 
    error_compress = ifelse(error < -0.75, -0.75, ifelse(error > 0.75, 0.75, error))
  ) %>% 
  ggplot() + 
  geom_sf(aes(col = error_compress), size = 0.4) + 
  theme_bw() + 
  scale_color_gradient2(
    low = "#ef8a62", mid = "#f7f7f7", high = "#67a9cf", 
    name = reslab,
    breaks = seq(from = -0.75, to = 0.75, by = 0.25),
    labels = c('< -0.75', '-0.5', '0.25', '0', '0.25', '0.5', '> 0.75')) +
  labs(x = "Meter", y = "Meter")
```

**SCBI**:

```{r}
scbi_fit_model <- focal_vs_comp_scbi %>% 
  fit_bayesian_model(model_specs = scbi_specs)
```

```{r}
scbi_growth_df <- focal_vs_comp_scbi %>% 
  predict_bayesian_model(model_specs = scbi_specs, posterior_param = scbi_fit_model) %>% 
  right_join(scbi_growth_df, by = c("focal_ID" = "stemID"))

scbi_growth_df <- scbi_growth_df %>% 
  st_as_sf()

# Observed vs predicted growth  
ggplot(scbi_growth_df, aes(x = growth, y = growth_hat)) +
  geom_point(size = 0.5, color = rgb(0, 0, 0, 0.25)) +
  stat_smooth(method = 'lm') +
  geom_abline(slope = 1, intercept = 0) +
  coord_fixed() + 
  labs(
    x = "Observed growth in dbh", y = "Predicted growth in dbh", 
    title = "Predicted vs Observed Growth"
  )

reslab <- expression(paste('Residual (cm ',y^{-1},')'))
scbi_growth_df %>% 
  # Need to investigate missingness
  filter(!is.na(growth_hat)) %>% 
  mutate(
    error = growth - growth_hat,
    error_bin = cut_number(error, n = 5), 
    error_compress = ifelse(error < -0.75, -0.75, ifelse(error > 0.75, 0.75, error))
  ) %>% 
  ggplot() + 
  geom_sf(aes(col = error_compress), size = 0.4) + 
  theme_bw() + 
  scale_color_gradient2(
    low = "#ef8a62", mid = "#f7f7f7", high = "#67a9cf", 
    name = reslab,
    breaks = seq(from = -0.75, to = 0.75, by = 0.25),
    labels = c('< -0.75', '-0.5', '0.25', '0', '0.25', '0.5', '> 0.75')) +
  labs(x = "Meter", y = "Meter")
```

### Run spatial cross-validation

For the above results we fit the model to the entire data set, and then make predictions across the entire data set from that fit. This could lead to overfitting because we are using the training data to also test the model. If model error is spatially correlated this could be a large issue (cite important sources here!). We can use the spatial block structure we defined above to deal with with. The function `run_cv` goes through each fold in the `cv_grid` and fits the model on all the other folds. Then applies that fit to the focal fold. It is a wrapper for `fit_bayesain_model` and `predict_bayesain_model` but fits a seperate model for each fold.

Here I cheated `run_cv` to run the cross valididation for just two training set. It runs on the whole data set up just for `test = fold 23` and `test = fold 2`. Just to see how long it takes.
```{r}
tic()

scbi_cv_predict <- focal_vs_comp_scbi %>%
  run_cv(model_specs = scbi_specs, max_dist = max_dist, cv_grid = scbi_cv_grid)
toc()
```

Running just two folds took 5440 seconds. There are 28 folds. So running everything should take 5440 * 14 / 60 / 60 = 21 hours.

Comapre results

```{r}

scbi_cv_predict %>%
  inner_join(scbi_growth_df, by = 'focal_ID', suffix = c('_cv','')) %>%
  summarise(rmse_cv = sqrt(mean((growth - growth_hat_cv)^2)),
            rmse = sqrt(mean((growth - growth_hat)^2)) )

```

### Run permutations

* Permutation test:
  + Add line to that shuffles competitor species within `focal_and_comp` just before fitting model
  + permutation AND CV: add `run_shuffle` argument to `run_cv()`

### Visualize posterior distributions

* Plot posterior distributions of all parameters

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# forestecology

[![Travis Build Status](https://travis-ci.org/rudeboybert/forestecology.svg?branch=master)](https://travis-ci.org/rudeboybert/forestecology)
[![CRAN status](https://www.r-pkg.org/badges/version/forestecology)](https://cran.r-project.org/package=forestecology)

@@

You can install the released version of `forestecology` from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("forestecology")
```

@@

``` r
# install.packages("devtools")
devtools::install_github("rudeboybert/forestecology")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```

What is special about using `README.Rmd` instead of just `README.md`? You can include R chunks like so:

```{r cars}
summary(cars)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub!
